---
title: "Data Viz"
author: "Qahhar"
date: "2023-12-30"
output:
  pdf_document: default
  html_document: default
---

```{r}

## load packages
library(janitor)
library(ggplot2)
library(dplyr)
library(lubridate)
library(zoo)
library(forcats)
library(ggthemes)
library(knitr)

```

# Part 1: Exploratory Data Analysis

The data we'll be using for this part of the project were downloaded from [kaggle](https://www.kaggle.com/aaronschlegel/seattle-pet-licenses/data), and include information about "pet licenses issued by the Seattle Animal Shelter between 2005 and early 2017." We'll be exploring these data and generating a few exploratory plots in the first part of the project.


## The Data

First, we'll read the data in from our `data/raw_data` directory.

```{r} 

# pets <- read.csv("/cloud/project/data/raw_data/seattle_pet_licenses.csv",
# stringsAsFactors = FALSE)

pets <- read.csv("/cloud/project/data/raw_data/seattle_pet_licenses.csv")

head(pets, n = 3)

```
It is noticed that a column `licence_issue_date` is not in the right format
```{r}
pets <- pets %>%
  mutate(license_issue_date = ymd_hms(license_issue_date)) %>%
  mutate(license_issue_date = format(license_issue_date, "%Y-%m-%d")) %>%
  arrange(desc(license_issue_date))

# pets <- pets %>%  
#   mutate(date =  lubridate::ymd_hms(license_issue_date),
#          ymd = as.yearmon(pets$date, "%y%m"))

head(pets, n = 3)
```

## Explore the Data

```{r}
#Understanding the structure 

glimpse(pets)

colnames(pets)

# #Summarizing factor/character columns only
# char_cols <- pets %>% select_if(is.character)
# summary(char_cols)
# 
# summary(pets$species)

```
- How many pet licenses are included in the dataset?

```{r}

sum(!is.na(pets$license_number))

```

- How many unique pet names are included in the dataset (animal_s_name)

```{r}
length(unique(pets$animal_s_name))

```

15796 unique names out of 66042. Find the more common names?

```{r}

# Use the pipe operator to find the most common names
ten_most_common_names <- pets %>%
  na.omit() %>%
  filter(animal_s_name != "") %>%
  group_by(animal_s_name) %>%
  summarize(count = n()) %>%
  arrange(desc(count)) %>%
  top_n(10)

# Display the result
head(ten_most_common_names, n = 3)
```


- How many different species are included in this dataset (species)?
```{r}
length(unique(pets$species))

```

- Which species are included in the dataset?
```{r}
unique(pets$species)

#Count of each unique specie
species_counts <- pets %>%
  group_by(species) %>%
  summarise(count = n()) 

print(species_counts)

```

## Visualize the Data

```{r} 
## visual breakdown of how many of each species 

colors <- c('#370B01', "#70402C", "#955732")
ggplot(data = pets) +
  geom_bar(aes(x=species, fill = species, fill = species))+ 
  scale_fill_manual(values = colors)
```

```{r} 
## Table: Most frequent Cat Name 
pets %>% 
  filter(species == "Cat", animal_s_name != "") %>%
  group_by(animal_s_name) %>% 
  summarise(n = n()) %>% 
  arrange(-n) %>%
  top_n(n = 10) %>%
  knitr::kable(., caption = "Top 10 Cat Names in Seattle")

#using either arrange(desc(n)) or arrange(-n) will achieve the same result:

# arrange(desc(n)): Sorts the values in descending order based on the column n.
# arrange(-n): Also sorts the values in descending order based on the column n.

```

```{r} 
## Table: Most frequent Dog Name
pets %>% 
  filter(species == "Dog", animal_s_name != "") %>%
  group_by(animal_s_name) %>% 
  summarise(n = n()) %>% 
  arrange(-n) %>%
  top_n(n = 10) %>%
  knitr::kable(., caption = "Top 10 Dog Names in Seattle")

```

- How the number of licenses recorded has changed over time?

```{r} 
# add date and ym columns
# pets$ym <- as.yearmon(pets$license_issue_date, "%y%m")

pets$ym <- as.yearmon(pets$license_issue_date, "%Y-%m-%d")

# how the number of licenses recorded has changed over time
pets %>%
  group_by(ym) %>% 
  ## count number within each group
  summarize(n= n()) %>%
  ggplot(., aes(ym, n)) + 
  ## geom name for line chart
  geom_line() +
  scale_x_yearmon() + 
  xlab("") + 
  ylab("Number of licenses")
```

# Part 2: Explanatory Data Analysis

The data used in this part of the project were dowloaded from [FiveThirtyEight](https://data.fivethirtyeight.com/) - `steak-survey`.  They were originally used in the article: [How Americans Like Their Steak](https://fivethirtyeight.com/features/how-americans-like-their-steak/). The goal of this part of the project will be to recreate the data visualization used in this article.

## The Data

The data focus on a question of choice,

Consider the following hypothetical situations 
- In Lottery A, you have a 50 chance of success with a payout of 100

- In Lottery B you have a 90 chance of success with a payout of 20

Assuming you have 10 to bet, would you play Lottery A or Lottery B.

```{r} 
## read in the data
steak <- read.csv("/cloud/project/data/raw_data/steak-risk-survey.csv") %>%
  ## remove first row which just includes the word "Response" in each column
  slice(2:n())

head(steak, n = 0)

```
Column names are lengthy, try changing?
```{r}
steak <- steak %>%
  rename(lottery_choice = Consider.the.following.hypothetical.situations...br.In.Lottery.A..you.have.a.50..chance.of.success..with.a.payout.of..100...br.In.Lottery.B..you.have.a.90..chance.of.success..with.a.payout.of..20...br..br.Assuming.you.have..10.to.bet..would.you.play.Lottery.A.or.Lottery.B.)

glimpse(steak)

```

## Explore the Data
- How many people participated in the survey?

```{r} 
nrow(steak)
```

- How many people responded “Yes” to the question “Do you eat steak?”
```{r} 

yes_steak <- steak %>%
  filter(Do.you.eat.steak. == 'Yes') %>%
  summarise(count = n())

yes_steak
```

- How many different (unique) responses were there to the question “How do you like your steak prepared?”

```{r} 

new_steak <- steak %>%
  filter(How.do.you.like.your.steak.prepared. != "") 
  
length(unique(new_steak$How.do.you.like.your.steak.prepared.))

unique(new_steak$How.do.you.like.your.steak.prepared.)
```

## Wrangle the Data

```{r} 

# Create a new data frame from steak data frame and assign it to the variable `pref`
pref <- steak %>% 
  #code below creates a column 'steak_pref' with its values being the unique characters of the How.do.you.like.your.steak.prepared. column from 'steak' being converted to factors
  mutate(steak_pref = factor(How.do.you.like.your.steak.prepared.,
                             levels = c("Well",
                                        "Medium Well",
                                        "Medium",
                                        "Medium rare",
                                        "Rare"))) %>% 
  #Filters out the values in the steak_pref column that are notated as ''
  filter(steak_pref != "") %>% 
  group_by(steak_pref) %>% 
  #Using the summarize function we create a new col 'n' which basically takes the count of individual groups
  summarise(n = n()) %>%
  # Create a new column that creates a ratio of each unique character, by dividing their sum over the total sum
  mutate(ratio = n / sum(n))

ncol(pref)
```

## Visualize the Data

```{r}
## generate the plot 
p <- ggplot(pref) +
  ## specify you want to generate a bar chart
  geom_bar(aes(x = steak_pref, y = ratio, fill = steak_pref),
                stat = 'identity', 
                width = 0.7) + 
  ## this adds text labels (you don't have to change anything here)
  geom_text(aes(label = paste0(as.integer(ratio*100),"%"),
                x = steak_pref,
                y = ratio), 
            stat= "identity", 
            hjust = -0.2, 
            size = 5,
            color = "grey40") +
  ## flip coordinates to make horizontal box plot
  coord_flip() +
  ## change the colors of the bars
  scale_fill_manual(values = c("#370B01", 
                           "#70402C",
                           "#955732", 
                           "#B36153",
                           "#B5413F")) +
  ## change the scale/labels of the steak-wellness variable (x-axis)
  scale_x_discrete(limits = levels(fct_rev(pref$steak_pref)),
                labels = c("Well", 
                           "Medium Well" = "Medium-well",
                           "Medium",
                           "Medium rare"="Medium-rare",
                           "Rare")) +
  # change the scale/labels of the percent axis (y-axis)
  scale_y_continuous(labels = scales::percent, 
                expand = c(mult = c(0,0), 
                           add = c(0,0.035))) +
  ## change the title, subtitle, and caption
  labs(title="'How Do You Like Your Steak Prepared?'", 
                subtitle="From a survey of 432 steak-eating Americans",
                caption="FiveThirtyEight: Surveymonkey") +
  ## change the theme (use ggthemes)
  theme_grey() +
  ## fine tune the theme
  theme(axis.text = element_text(size = 14), 
                title = element_text(size = 16), 
                legend.position="none", 
                plot.caption=element_text(size = 12), 
                panel.grid.major = element_blank(), 
                panel.grid.minor = element_blank(),
                axis.text.x = element_blank() 
  )

p
```

## Save the Plot

```{r}
## save plot to figures/explanatory_figures directory
ggsave(plot = p, filename = '/cloud/project/steak_graph.png', width = 8, height = 4)
```



## Session Info 
This tells people what software versions you were using when you ran this notebook. 

```{r}
sessionInfo()
```